\phantomsection  % so hyperref links will lead to the correct place in document
\addcontentsline{toc}{chapter}{Acknowledgments}

\chapter*{Acknowledgments}
\label{chapter:acknowledgments}

I would like express my sincerest gratitude to my supervisor, Henryk Maciejewski, for~the~whole his help, not only during my research and redaction of this dissertation, but~also, most importantly, for his assistance in introducing me the University ecosystem and guiding me how to survive in it. After all our countless conversations, I~understand how to continue doing my part well, while still remaining a good and a kind person.

I would like to thank my fiancée, Sylwia, for invaluable support, unlimited patience and especially for believing in~me and not asking about my daily writing progress~;-) I~hope that now we will be able to fully focus on our future adventures together.

I would also like to send my thanks to everyone who made this whole period bearable and kept me sanity during this journey – most importantly: Adrian, Arie, Artur, Arx, Bogna, Daniel, David, Dawid, Ignacio, Karolina, Klaudia, Mateusz, Mike and Piotr. Without You, my motivation would have been lost long ago...

Finally, I would like to express my deep admiration and respect to Alexandra Asanovna Elbakyan, who has enabled free access to the scientific publications for countless number of researchers and scientists around the world, making the science open and not closed behind paywalls.

\noindent
To all of You, from the deepest bottom of my heart, \\
Thank You so much! / Dziękuję bardzo! / Tack så mycket!

\hfill – Szymon Datko
